---
title: "Deterministic vs. stochastic models"
author: "Janetta Skarp"
date: "10 November 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction

* Using stochastic models (especially individual-based models) is currently trendy
* Stochastic models can be more difficult to deal with than deterministic models
* For some questions, it may not matter whether you use a stochastic or deterministic model.
* When is using a stochastic model a waste of your time?

## Modelling

Simple SIR models with individuals moving from susceptible to infected and from infected to recovered (no births or deaths) were constructed.

## Deterministic model

A deterministic SIR model was constructed:

```{r deterministic_model, echo = FALSE, cache = TRUE}
library("deSolve") #package for solving differential equations

## Make an SIS function
sir <- function(time, state, parameters) {
  
  # define model parameters in term of the natural parameters
  beta <- parameters["R0"]/parameters["D_inf"] 
  gamma <- 1/parameters["D_inf"]
  
  with(as.list(c(state, parameters)), {
    
    dS <- -beta * S * I 
    dI <-  beta * S * I - gamma * I
    dR <-  gamma * I
    
    return(list(c(dS, dI, dR)))
  })
}

#############################
### THINGS YOU CAN CHANGE ###
#############################

# Proportion in each compartment at the start
init  <- c(
  S = 1-1e-4, 
  I = 1e-4,
  R = 0
) # N = 1

## R0 = basic reproduction number, D_inf = duration of infection
parameters <- c(
  R0 = 3, 
  D_inf = 2
)

## Timeframe
times <- seq(0, 100, by = 1)

## Solve using General Solver for Ordinary Differential Equations (ode)
run <- ode(y = init, times = times, func = sir, parms = parameters)
run_det <- as.data.frame(run) # change to data frame
run_det$time <- NULL # Delete time variable

## Plot model
par(mfrow=c(1,1))

matplot(x = times, y = run_det, type = "l",
  xlab = "Time", ylab = "Proportion susceptible or infected",  main = "Deterministic SIR Model",
  lwd = 1, lty = 1, bty = "l", col = c("black","red","orange"))

## Add legend
legend(80, 0.8, c("Susceptible", "Infected", "Recovered"), pch = 1, col = c("black","red","orange"), bty = "n")
```

## Stochastic model

A stochastic model was constructed: 

```{r stochastic_model, echo = FALSE, warning = FALSE, cache = TRUE}
##########################
## Input values for SIR ##
##########################

# Time
timestep <- 0.1
times <- seq(0, 100, by = timestep)

# Initial population: N-1 susceptible, 1 infectious, 0 recovered
init.values = c(
  S = 100-1,
  I = 1,
  R = 0
)
N = sum(init.values)

# Invent some epidemic data to fit my model to
true_infectious <- (dnorm(times, 20, 4))*N
#plot(true_infectious)

###############
## The model ##
###############

# Array for holding collective disease status information for whole period of time
data <- array(0, dim =c(length(times), length(init.values)+3))
data[,1] <- times # make first column the timesteps to make plotting easier later on
  
# R0 & duration of infectiousness
R0 <- 3
D_inf <- 2
  
# Calculating probabilities that do not change with time
p <- R0 * (timestep/(D_inf*N)) # probability of effective contact
r_t <- (1/D_inf)*timestep # Recovery rate
  
# For loops for calculating the numbers susceptible, infected, and recovered at each timepoint
for (time in times){
  if (time == 0){ # Set up the number of S/I/R at time 0
    data[1,2] <- init.values["S"] # number of susceptibles at time 0
    data[1,3] <- init.values["I"] # number of infecteds at time 0
    data[1,4] <- init.values["R"] # number of recovereds at time 0
    data[1,5] <- init.values["I"] # number newly infected at time 0
    data[1,6] <- init.values["R"] # number newly recovered at time 0
      
  } else{
    whole_time <- 1/timestep * time # makes time into the whole number that it corresponds to in the array
    foi_t <- 1 - (1 - p)^data[whole_time, 3] # Force of infection at time t, affected by the number infectious at prev t
    inf <- rbinom(1, size = data[whole_time,2], prob = foi_t) # number who become infected in this timestep
    rec <- rbinom(1, size = data[whole_time,3], prob = r_t)# number who become recovered in this timestep
      
    data[whole_time+1,2] <- data[whole_time,2] - inf # number of susceptibles at other times
      
    data[whole_time+1,3] <- data[whole_time,3]  + inf - rec # number of infecteds at other times
      
    data[whole_time+1,4] <- data[whole_time,4] + rec # number of recovereds at other times
    
    data[whole_time+1,5] <- data[whole_time+1,3] - data[whole_time,3] + data[whole_time+1,4] - data[whole_time,4] # number of newly infected
    
    data[whole_time+1,6] <- data[whole_time+1,4] - data[whole_time,4] # number of newly recovered
  }
}

###############
## SIR plots ##
###############

run_stoch <- data.frame(data) # make array into a dataframe
colnames(run_stoch) <- c("time","S", "I", "R", "new_I", "new_R")

par(mfrow = c(1,1))

# Plot for SIR model
plot(x = run_stoch$time, y = run_stoch$I, type = "line", col = "red", ylim = c(0,N),
     xlab = "Time", ylab = "Number susceptible/infected/recovered", main = "Stochastic SIR Model")
par(new=T)
plot(x = run_stoch$time, y = run_stoch$S, type = "line", ylim = c(0,N), ylab = "", xlab = "") # add susceptible line
par(new=T)
plot(x = run_stoch$time, y = run_stoch$R, type = "line", col = "orange", ylim = c(0,N), ylab = "", xlab = "") # recovered

# Add legend
legend(60, 0.8*N, c("Susceptible", "Infected", "Recovered"), pch = 1, col = c("black", "red", "orange"), bty = "n")

# Plot for newly infected and newly recovered
#plot(x = run_stoch$time, y = run_stoch$new_I, type = "line", col = "red", ylim = c(0,max(run_stoch$new_I)),
#     xlab = "Time", ylab = "Number newly infected/recovered", main = "New infections/recoveries")
#par(new=T)
#plot(x = run_stoch$time, y = run_stoch$new_R, type = "line", col = "orange", ylim = c(0,max(run_stoch$new_I)), ylab = "", xlab = "") # add susceptible line

```

## Markov Chain Monte Carlo

A Metropolis-Hastings algorithm was developed to infer the values of beta, the rate at which susceptibles become infectious, and gamma, the rate at which infectious individuals recover, in the stochastic SIR model.
